---
title: "Microarray Data Analysis"
author: "Jialin Ma"
date: "October 17, 2018"
output:
  html_document:
    self_contained: false
    toc: true
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE)
```

## Load Packages

```{r}
suppressPackageStartupMessages({
    library(here)
    library(oligo)
})
```


## Load the dataset

```{r}
dataset <- read.celfiles(list.files(here("data/microarray"), full.names = TRUE))
dataset <- dataset[, c(paste0(c("GSM651310", "GSM651320", "GSM651315", "GSM651325"), ".CEL.gz"))]
sampleNames(dataset) <- c("control_rep1", "control_rep2", "treat_rep1", "treat_rep2")
pData(dataset)$group <- c("control", "control", "treat", "treat")
dataset
```

## Quality control

### Boxplot

```{r}
boxplot(dataset, target = "core")
```


### MA plot

```{r}
oligo::MAplot(dataset, pairs = TRUE, ylim = c(-1, 1))
```


### Quality report of the raw data

The following will generate a quality report of the raw microarray data in 
`docs/microarray_qualitymetrics` directory.

```r
library(arrayQualityMetrics)
arrayQualityMetrics(expressionset = dataset,
    outdir = "docs/microarray_qualitymetrics",
    force = TRUE, do.logtransform = TRUE,
    intgroup = c("group"))
```

## RMA

The RMA method proceeds with background subtraction, normalization and summarization
using a deconvolution method for background correction, quantile normalization and
the RMA (robust multichip average) algorithm for summarization.

```{r}
edata <- oligo::rma(dataset)
edata
```

## Access the quality after normalization

### PCA plot before and after normalization

```{r}
library(ggplot2)
library(cowplot)
plot_grid(
    local({
        PCA_raw <- prcomp(t(log2(exprs(dataset))), scale = FALSE)
    
        dataGG <- data.frame(PC1 = PCA_raw$x[,1], PC2 = PCA_raw$x[,2],
                         Group = pData(dataset)$group)
    		
        qplot(PC1, PC2, data = dataGG, color = Group,
              main = "PCA plot of the raw data (log-transformed)", asp = 1.0, geom = "text",
    	label = sampleNames(dataset)) + scale_colour_brewer(palette = "Set2")
    }),
    local({
        PCA <- prcomp(t(exprs(edata)), scale = FALSE)
        dataGG <- data.frame(PC1 = PCA$x[,1], PC2 = PCA$x[,2],
                  Group = pData(dataset)$group)
        qplot(PC1, PC2, data = dataGG, color =  Group,
            main = "PCA plot of the normalized data", asp = 1.0, geom = "text",
            label = sampleNames(edata)) +
            scale_colour_brewer(palette = "Set2")
    })
)
```

### Boxplot before and after normalization

```{r}
par(mfrow = c(1,2))
boxplot(dataset)
boxplot(edata)
par(mfrow = c(1,1))
```

### MA plot after normalization

```{r}
oligo::MAplot(edata, pairs = TRUE, ylim = c(-1, 1))
```

### Heatmap with sample-to-sample distance after normalization

It can not provide too much information for us since the number of samples is limited.

```{r}
library(RColorBrewer)
library(pheatmap)
dists <- as.matrix(dist(t(exprs(edata)), method = "manhattan"))
diag(dists) <- NA
hmcol <- colorRampPalette(rev(brewer.pal(9, "PuOr")))(255)
pheatmap(dists, col = rev(hmcol), clustering_distance_rows = "manhattan",
				    clustering_distance_cols = "manhattan")
```

## Filter based on intensity

Microarray data commonly show a large number of probes in the background
intensity range. They also do not change much across arrays. Hence they
combine a low variance with a low intensity. We want to filter these results
as they may contribute to false positive results in the differential expression
analysis.

```{r}
edata_medians <- rowMedians(exprs(edata))
hist_res <- hist(edata_medians, 100, col="#e7efd8", freq = FALSE,
                 main = "Histogram of the median intensities",
                 xlab = "Median intensities")
emp_mu <- hist_res$breaks[which.max(hist_res$density)]
emp_sd <- mad(edata_medians)/2
prop_cental <- 0.50
lines(sort(edata_medians),
      prop_cental*dnorm(sort(edata_medians), mean = emp_mu, sd = emp_sd),
    col = "grey10", lwd = 4)

cut_val <- 0.05 / prop_cental
thresh_median <- qnorm(0.05 / prop_cental, emp_mu, emp_sd)
samples_cutoff <- 2
idx_thresh_median <- apply(exprs(edata), 1, function(x){
				   sum(x > thresh_median) >= samples_cutoff})
table(idx_thresh_median)
edata <- subset(edata, idx_thresh_median)
```


## Identification of differentially expressed genes

Create a simple design matrix.

```{r}
library(limma)
f <- factor(c("control", "control", "treat", "treat"))
design <- model.matrix(~ 0 + f)
colnames(design)
colnames(design) <- c("control", "treat")
design
```

We can fit the linear model, define appropriate contrast to test the hypothesis on
treatment effect and compute the moderated t–statistics by calling
the `eBayes` function.

```{r}
data.fit <- lmFit(exprs(edata), design)
head(data.fit$coefficients)
contrast.matrix <- makeContrasts(treat-control,levels=design)
data.fit.con <- contrasts.fit(data.fit,contrast.matrix)
data.fit.eb <- eBayes(data.fit.con)
```

Volcano plot to show the distribution of fold change and p value.

```{r}
volcanoplot(data.fit.eb,highlight=10)
```

Then we sort the results by their absolute t-statistics.

```{r}
top <- topTable(data.fit.eb, number = Inf)
head(top)
```

Check how many results can we get if we use a p value cutoff by
0.001 or an adjusted p value cutoff by 0.05.

```{r}
table(top$adj.P.Val < 0.05)
table(top$P.Value < 0.001)
```

We notice that the number of differential expressed genes are limited, which indicates the
the effect of 4 hour DEX treatment may not be significant in altering the gene expression.

Then plot a histogram of p value distribution.

```{r}
hist(top$P.Value, col = brewer.pal(3, name = "Set2")[1],
      main = "Control vs Treatment of DEX by 4 hours", xlab = "p-values")
```

We zoom in the plot to see p value distribution between 0 and 0.1.

```{r}
hist(top$P.Value[top$P.Value < 0.1], col = brewer.pal(3, name = "Set2")[1],
      main = "Control vs Treatment of DEX by 4 hours", xlab = "p-values")
```

Zoom in again and the orange bin indicates the differentially expressed genes
if we use a p value cutoff by 0.001.

```{r}
local({
    breaks <- 0:10/1000
    color1 <- brewer.pal(3, name = "Set2")[1]
    color2 <- brewer.pal(3, name = "Set2")[2]
    color <- rep.int(color1, length(breaks) - 1)
    color[1] <- color2
    data <- top$P.Value[top$P.Value < 0.01]
    hist(data, col = color,
         main = "Control vs Treatment of DEX by 4 hours", xlab = "p-values", breaks = breaks)
})
```


## Annotating genes

We need to annotate the gene names by the probe IDs.

```{r}
suppressPackageStartupMessages({
    library(hgu95av2.db)
})
get_symbol <- function(probeid) {
    ans <- mapIds(hgu95av2.db::hgu95av2.db,
                  keys = probeid, keytype = "PROBEID", column = "SYMBOL", multiVals = "first")
    unname(ans)
}
get_genename <- function(probeid) {
    ans <- mapIds(hgu95av2.db::hgu95av2.db,
                  keys = probeid, keytype = "PROBEID", column = "GENENAME", multiVals = "first")
    unname(ans)
}
top$symbol <- get_symbol(rownames(top))
top$gene_name <- get_genename(rownames(top))
```

## Generate a table of differentially expressed genes

We use a p value cutoff by 0.001.

```{r}
selected <- top[, c("symbol", "gene_name", "logFC", "P.Value", "adj.P.Val")]
#top <- cbind(data.frame(probeID = rownames(top), stringsAsFactors = FALSE), top)
#rownames(top) <- NULL
selected <- selected[selected$P.Value < 0.001,]
knitr::kable(selected)
```


## Gene clustering

The following is a heatmap of gene expression values of the differentially
expressed genes.

```{r}
library(pheatmap)
mat <- exprs(edata)[rownames(selected), ]
rownames(mat) <- selected$symbol
pData(edata)$sample_name <- rownames(pData(edata))
anno <- as.data.frame(pData(edata))[, c("group", "sample_name")]
pheatmap(mat, annotation_col = anno,
         main = "Heatmap of gene expression levels")
```

We can roughly divide the genes (probes) into two clusters, and the control/treat samples
are also clearly separated.

## PCA plot


```{r}
library(ggfortify)
autoplot(prcomp(exprs(edata)[rownames(selected), ]))
```

The PCA plot did not provide much useful information.


