---
title: "Microarray Data Analysis"
author: "Jialin Ma"
date: "October 17, 2018"
output:
  html_document:
    self_contained: false
    toc: true
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE)
```

## Load Packages

```{r}
suppressPackageStartupMessages({
    library(here)
    library(oligo)
})
```


## Load the dataset

- GSM651311 Keratinocytes, untreated 24h, rep1
- GSM651316 Keratinocytes, DEX-treated 24h, rep1
- GSM651321 Keratinocytes, untreated 24h, rep2
- GSM651326 Keratinocytes, DEX-treated 24h, rep2

```{r}
dataset <- read.celfiles(list.files(here("data/microarray"), full.names = TRUE))
#dataset <- dataset[, c(paste0(c("GSM651310", "GSM651320", "GSM651315", "GSM651325"), ".CEL.gz"))]
dataset <- dataset[, c(paste0(c("GSM651311", "GSM651321", "GSM651316", "GSM651326"), ".CEL.gz"))]
sampleNames(dataset) <- c("control_rep1", "control_rep2", "treat_rep1", "treat_rep2")
pData(dataset)$group <- c("control", "control", "treat", "treat")
dataset
```

## Quality control

### Boxplot

```{r}
boxplot(dataset, target = "core")
```

### MA plot

```{r}
oligo::MAplot(dataset, pairs = TRUE, ylim = c(-1, 1))
```


### Quality report of the raw data

The following will generate a quality report of the raw microarray data in 
`docs/microarray_qualitymetrics` directory.

```r
library(arrayQualityMetrics)
arrayQualityMetrics(expressionset = dataset,
    outdir = "docs/microarray_qualitymetrics",
    force = TRUE, do.logtransform = TRUE,
    intgroup = c("group"))
```

## RMA

The RMA method proceeds with background subtraction, normalization and summarization
using a deconvolution method for background correction, quantile normalization and
the RMA (robust multichip average) algorithm for summarization.

```{r}
edata <- oligo::rma(dataset)
edata
```

## Access the quality after normalization

### PCA plot before and after normalization

```{r}
library(ggplot2)
library(cowplot)
plot_grid(
    local({
        PCA_raw <- prcomp(t(log2(exprs(dataset))), scale = FALSE)
    
        dataGG <- data.frame(PC1 = PCA_raw$x[,1], PC2 = PCA_raw$x[,2],
                         Group = pData(dataset)$group)
    		
        qplot(PC1, PC2, data = dataGG, color = Group,
              main = "PCA plot of the raw data (log-transformed)", asp = 1.0, geom = "text",
    	label = sampleNames(dataset)) + scale_colour_brewer(palette = "Set2")
    }),
    local({
        PCA <- prcomp(t(exprs(edata)), scale = FALSE)
        dataGG <- data.frame(PC1 = PCA$x[,1], PC2 = PCA$x[,2],
                  Group = pData(dataset)$group)
        qplot(PC1, PC2, data = dataGG, color =  Group,
            main = "PCA plot of the normalized data", asp = 1.0, geom = "text",
            label = sampleNames(edata)) +
            scale_colour_brewer(palette = "Set2")
    })
)
```

### Boxplot before and after normalization

```{r}
par(mfrow = c(1,2))
boxplot(dataset)
boxplot(edata)
par(mfrow = c(1,1))
```

### MA plot after normalization

```{r}
oligo::MAplot(edata, pairs = TRUE, ylim = c(-1, 1))
```

### Heatmap with sample-to-sample distance after normalization

It can not provide too much information for us since the number of samples is limited.

```{r}
library(RColorBrewer)
library(pheatmap)
dists <- as.matrix(dist(t(exprs(edata)), method = "manhattan"))
diag(dists) <- NA
hmcol <- colorRampPalette(rev(brewer.pal(9, "PuOr")))(255)
pheatmap(dists, col = rev(hmcol), clustering_distance_rows = "manhattan",
				    clustering_distance_cols = "manhattan")
```

## Filter based on intensity

Microarray data commonly show a large number of probes in the background
intensity range. They also do not change much across arrays. Hence they
combine a low variance with a low intensity. We want to filter these results
as they may contribute to false positive results in the differential expression
analysis. The bars represent the distribution of median intensities, the red
vertical line represents the threshold value. If more than two samples of a
gene have intensities larger than the threshold value, the gene will be
kept.

```{r}
edata_medians <- rowMedians(exprs(edata))
hist_res <- hist(edata_medians, 100, col="#e7efd8", freq = FALSE,
                 main = "Histogram of the median intensities",
                 xlab = "Median intensities")
emp_mu <- hist_res$breaks[which.max(hist_res$density)]
emp_sd <- mad(edata_medians)/2
prop_cental <- 0.50
lines(sort(edata_medians),
      prop_cental*dnorm(sort(edata_medians), mean = emp_mu, sd = emp_sd),
    col = "grey10", lwd = 3)

#cut_val <- 0.05 / prop_cental
thresh_median <- qnorm(0.05 / prop_cental, emp_mu, emp_sd)
abline(v = thresh_median, lwd = 4, col = "red")

samples_cutoff <- 2
idx_thresh_median <- apply(exprs(edata), 1, function(x){
				   sum(x > thresh_median) >= samples_cutoff})
table(idx_thresh_median)
edata <- subset(edata, idx_thresh_median)
```


## Identification of differentially expressed genes

Create a design matrix. We will also consider the batch effects between replicates
in order to remove them.

```{r}
library(limma)
f <- factor(c("control", "control", "treat", "treat"))
batch <- factor(c("rep1", "rep2", "rep1", "rep2"))
design <- model.matrix(~ 0 + f + batch)
colnames(design)
colnames(design) <- c("control", "treat", "batch")
design
```

We can fit the linear model, define appropriate contrast to test the hypothesis on
treatment effect and compute the moderated t–statistics by calling
the `eBayes` function.

```{r}
data.fit <- lmFit(exprs(edata), design)
head(data.fit$coefficients)
contrast.matrix <- makeContrasts(treat-control,levels=design)
data.fit.con <- contrasts.fit(data.fit,contrast.matrix)
data.fit.eb <- eBayes(data.fit.con)
```

Volcano plot to show the distribution of fold change and p value.

```{r}
volcanoplot(data.fit.eb,highlight=10)
```

Then we sort the results by their absolute t-statistics.

```{r}
top <- topTable(data.fit.eb, number = Inf)
head(top)
```

Check how many results can we get if we use a p value cutoff by
0.001 or an adjusted p value cutoff by 0.05.

```{r}
table(top$adj.P.Val < 0.05)
table(top$P.Value < 0.001)
```

We would like to visualize the distribution of p value with a histogram.

```{r}
hist(top$P.Value, col = brewer.pal(3, name = "Set2")[1], breaks = 100,
      main = "Control vs Treatment of DEX", xlab = "p-values")
```


## Annotating genes

We need to annotate the gene names by the probe IDs.

```{r}
suppressPackageStartupMessages({
    library(hgu95av2.db)
})
get_symbol <- function(probeid) {
    ans <- mapIds(hgu95av2.db::hgu95av2.db,
                  keys = probeid, keytype = "PROBEID", column = "SYMBOL", multiVals = "first")
    unname(ans)
}
get_genename <- function(probeid) {
    ans <- mapIds(hgu95av2.db::hgu95av2.db,
                  keys = probeid, keytype = "PROBEID", column = "GENENAME", multiVals = "first")
    unname(ans)
}
top$symbol <- get_symbol(rownames(top))
top$gene_name <- get_genename(rownames(top))
```

## Generate a table of differentially expressed genes

We will use a cutoff of adjusted p value by 0.05.

```{r}
selected <- top[, c("symbol", "gene_name", "logFC", "P.Value", "adj.P.Val")]
#top <- cbind(data.frame(probeID = rownames(top), stringsAsFactors = FALSE), top)
#rownames(top) <- NULL
selected <- selected[selected$adj.P.Val < 0.05,]
nrow(selected)
```

We will only show the first 100 genes.

```{r}
knitr::kable(selected[1:100,])
```


## Gene clustering

The following is a heatmap of gene expression values of the differentially
expressed genes.

```{r}
library(pheatmap)
mat <- exprs(edata)[rownames(selected), ]
rownames(mat) <- selected$symbol
pData(edata)$sample_name <- rownames(pData(edata))
anno <- as.data.frame(pData(edata))[, c("group", "sample_name")]
pheatmap(mat, annotation_col = anno, show_rownames = FALSE,
         main = "Heatmap of gene expression levels")
```

We can roughly divide the genes (probes) into three clusters, and the control/treat samples
are also clearly separated.

## PCA plot


```{r}
library(ggfortify)
autoplot(prcomp(exprs(edata)[rownames(selected), ]))
```

The PCA plot did not provide much useful information.


## Pathway analysis

```{r}
library(ReactomePA)
entrez_ids <- mapIds(hgu95av2.db::hgu95av2.db, keys = rownames(selected),
                     keytype = "PROBEID", column = "ENTREZID", multiVals = "first")
entrez_ids <- entrez_ids[!is.na(entrez_ids)]
reactome_enrich <- enrichPathway(gene = entrez_ids, organism = "human")
```

```{r fig.width=15, fig.height=10}
barplot(reactome_enrich, showCategory = Inf)
dotplot(reactome_enrich, showCategory = Inf)
```

```{r fig.width=12, fig.height=10}
ReactomePA::emapplot(reactome_enrich, showCategory = Inf)
```


